Fast and scalable Gaussian process modeling with applications to astronomical time series \vspace{-3\baselineskip}

### 2017-01-15 paper/ms.tex
typo in extended matrix
words about implementation

@@

To provide some insight for this method, we will start by working through the
simplest possible example.
In this case, we'll assume that we have three data points
$\{y_1,\,y_2,\,y_3\}$ observed at times $\{t_1,\,t_2,\,t_3\}$ with
uncertainties $\{{\sigma_1}^2,\,{\sigma_2}^2,\,{\sigma_3}^2\}$ and we would
like to compute the likelihood of these data under a GP model with the
covariance function
\begin{eqnarray}
k(\tau_{nm}) = \sigma_n^2\,\delta_{nm} + a\,\exp(-c\,\tau_{nm})\quad.
\end{eqnarray}
This function is identical to \eq{celerite-kernel-complex} with $J=1$, $b=0$,
and $d=0$.
To demonstrate motivate this method, we will write out the full system of
equations that we must solve to apply the inverse of $K$.
In matrix notation, this can be written as
\begin{eqnarray}
K\,\bvec{x} &=& \bvec{y} \\
\begin{pmatrix}
    a+{\sigma_1}^2 & a\,e^{-c\,\tau_{2,1}} & a\,e^{-c\,\tau_{3,1}}\\
    a\,e^{-c\,\tau_{2,1}} & a+{\sigma_2}^2 & a\,e^{-c\,\tau_{3,2}}\\
    a\,e^{-c\,\tau_{3,1}} & a\,e^{-c\,\tau_{3,2}} & a+{\sigma_3}^2
\end{pmatrix}\,
\begin{pmatrix}
    x_1 \\ x_2 \\ x_3
\end{pmatrix} &=&
\begin{pmatrix}
    y_1 \\ y_2 \\ y_3
\end{pmatrix} \quad.
\end{eqnarray}
Now, if we introduce the variables
\begin{eqnarray}
    u_n = e^{-c\,\tau_{n+2,n+1}}\,u_{n+1} + a\,x_{n+1}
\end{eqnarray}
where $u_{N} = 0$, and
\begin{eqnarray}\eqlabel{algo-first}
    g_n = e^{-c\,\tau_{n+1,n}}\,g_{n-1} + e^{-c\,\tau_{n+1,n}}\,x_{n}
\end{eqnarray}
where $g_{0} = 0$,
the system of equations can be rewritten as
\begin{eqnarray}
(a+{\sigma_1}^2)\,x_1 + e^{c\,\tau_{2,1}}\,u_1 &=& y_1 \\
a\,g_1 + (a+{\sigma_2}^2)\,x_2 + e^{c\,t_2}\,u_2 &=& y_2 \\
a\,g_2 + (a+{\sigma_3}^2)\,x_3 &=& y_3 \quad. \eqlabel{algo-last}
\end{eqnarray}
The system defined by \eq{algo-first} through \eq{algo-last} can be
rewritten as a matrix equation to show the benefit that this seemingly trivial
reformulation provides:
\begin{eqnarray}
\begin{pmatrix}
    a+{\sigma_1}^2 & e^{c\,t_1} & 0 & 0 & 0 & 0 & 0 \\
    e^{c\,t_1} & 0 & -1 & 0 & 0 & 0 & 0 \\
    0 & -1 & 0 & e^{-c\,t_2} & a & 0 & 0 \\
    0 & 0 & e^{-c\,t_2} & a+{\sigma_2}^2 & e^{c\,t_2} & 0 & 0 \\
    0 & 0 & a & e^{c\,t_2} & 0 & -1 & 0 \\
    0 & 0 & 0 & 0 & -1 & 0 & e^{-c\,t_3} \\
    0 & 0 & 0 & 0 & 0 & e^{-c\,t_3} & a+{\sigma_3}^2 \\
\end{pmatrix}\,
\begin{pmatrix}
    x_1 \\ u_1 \\ g_1 \\ x_2 \\ u_2 \\ g_2 \\ x_3
\end{pmatrix} &=&
\begin{pmatrix}
    y_1 \\ 0 \\ 0 \\ y_2 \\ 0 \\ 0 \\ y_3
\end{pmatrix} \quad.
\end{eqnarray}

>>>>

\subsection{An example}

To provide some insight for this method, we will start by working through the
simplest possible example.
In this case, we'll assume that we have three data points
$\{y_1,\,y_2,\,y_3\}$ observed at times $\{t_1,\,t_2,\,t_3\}$ with
uncertainties $\{{\sigma_1}^2,\,{\sigma_2}^2,\,{\sigma_3}^2\}$ and we would
like to compute the likelihood of these data under a GP model with the
covariance function
\begin{eqnarray}
k(\tau_{nm}) = \sigma_n^2\,\delta_{nm} + a\,\exp(-c\,\tau_{nm})\quad.
\end{eqnarray}
This function is identical to \eq{celerite-kernel-complex} with $J=1$, $b=0$,
and $d=0$.
To demonstrate motivate this method, we will write out the full system of
equations that we must solve to apply the inverse of $K$.
In matrix notation, this can be written as
\begin{eqnarray}\eqlabel{impl-matrix}
K\,\bvec{x} &=& \bvec{y} \\
\begin{pmatrix}
    a+{\sigma_1}^2 & a\,e^{-c\,\tau_{2,1}} & a\,e^{-c\,\tau_{3,1}}\\
    a\,e^{-c\,\tau_{2,1}} & a+{\sigma_2}^2 & a\,e^{-c\,\tau_{3,2}}\\
    a\,e^{-c\,\tau_{3,1}} & a\,e^{-c\,\tau_{3,2}} & a+{\sigma_3}^2
\end{pmatrix}\,
\begin{pmatrix}
    x_1 \\ x_2 \\ x_3
\end{pmatrix} &=&
\begin{pmatrix}
    y_1 \\ y_2 \\ y_3
\end{pmatrix} \quad.
\end{eqnarray}
Now, if we introduce the variables
\begin{eqnarray}
    u_n = e^{-c\,\tau_{n+2,n+1}}\,u_{n+1} + a\,x_{n+1}
\end{eqnarray}
where $u_{N} = 0$, and
\begin{eqnarray}\eqlabel{algo-first}
    g_n = e^{-c\,\tau_{n+1,n}}\,g_{n-1} + e^{-c\,\tau_{n+1,n}}\,x_{n}
\end{eqnarray}
where $g_{0} = 0$,
the system of equations can be rewritten as
\begin{eqnarray}
(a+{\sigma_1}^2)\,x_1 + e^{-c\,\tau_{2,1}}\,u_1 &=& y_1 \\
a\,g_1 + (a+{\sigma_2}^2)\,x_2 + e^{-c\,\tau_{3,2}}\,u_2 &=& y_2 \\
a\,g_2 + (a+{\sigma_3}^2)\,x_3 &=& y_3 \quad. \eqlabel{algo-last}
\end{eqnarray}
The system defined by \eq{algo-first} through \eq{algo-last} can be
rewritten as a matrix equation to show the benefit that this seemingly trivial
reformulation provides:
\begin{eqnarray}
\begin{pmatrix}
    a+{\sigma_1}^2 & e^{-c\,\tau_{2,1}} & 0 & 0 & 0 & 0 & 0 \\
    e^{-c\,\tau_{2,1}} & 0 & -1 & 0 & 0 & 0 & 0 \\
    0 & -1 & 0 & a & e^{-c\,\tau_{3,2}} & 0 & 0 \\
    0 & 0 & a & a+{\sigma_2}^2 & e^{-c\,\tau_{3,2}} & 0 & 0 \\
    0 & 0 & e^{-c\,\tau_{3,2}} & e^{-c\,\tau_{3,2}} & 0 & -1 & 0 \\
    0 & 0 & 0 & 0 & -1 & 0 & a \\
    0 & 0 & 0 & 0 & 0 & a & a+{\sigma_3}^2 \\
\end{pmatrix}\,
\begin{pmatrix}
    x_1 \\ u_1 \\ g_1 \\ x_2 \\ u_2 \\ g_2 \\ x_3
\end{pmatrix} &=&
\begin{pmatrix}
    y_1 \\ 0 \\ 0 \\ y_2 \\ 0 \\ 0 \\ y_3
\end{pmatrix}\nonumber
\end{eqnarray}
where we will follow \citet{Ambikasaran:2015} and call this the ``extended''
system and rewrite \eq{impl-matrix} as
\begin{eqnarray}
    K_\mathrm{ext}\,\bvec{x}_\mathrm{ext} &=& \bvec{y}_\mathrm{ext} \quad.
\end{eqnarray}
Even though $K_\mathrm{ext}$ is a larger matrix than the original $K$ that we
started with, it is now has a sparse banded structure that can be exploited to
solve the system efficiently.
In particular, band matrix solvers are available that can perform an
LU-decomposition of banded matrices like this in $\mathcal{O}(N)$
operations~--~instead of the $\mathcal{O}(N^3)$ that would be required in
general~--~and we can take advantage of these algorithms to solve our system
exactly because the target vector $\bvec{x}$ is a subset of the elements of
$\bvec{x}_\mathrm{ext}$.

In the following section, we will discuss this method more generally but it's
worth noting a few important facts that can already be seen in this example.
First, the fundamental reason why this matrix $K$ can be solved efficiently is
the following property of exponentials
\begin{eqnarray}
    e^{-c\,(t_3 - t_2)} \, e^{-c\,(t_2 - t_1)} =
    e^{-c\,(t_3 - t_2 + t_2 - t_1)} =
    e^{-c\,(t_3 - t_1)}
\end{eqnarray}
and it is important to note that this property does not extend to other common
covariance functions like the ``exponential-squared'' function
\begin{eqnarray}
    k(\tau) \propto e^{-c\,\tau^2} \quad.
\end{eqnarray}
Furthermore, our derivation of the extended matrix requires that the data
points be monotonically sorted in time.
Neither of these properties will be satisfied in general for multidimensional
inputs and all of our following discussion will assume a sorted
one-dimensional dataset.

@@

\citet{Ambikasaran:2015} demonstrated two key facts that allow us to use this
extended matrix formalism in practice.
First, even if the covariance function is a mixture of exponentials, the
extended matrix will still be banded with bandwidth \dfmtodo{check}.
Second, \citet{Ambikasaran:2015} proved that the determinant of
$K_\mathrm{ext}$ is equal to the determinant of $K$ up to a sign.
